Nested Sampling with Multiple Objective Functions

### 2018-06-15 doc/ms.tex
More writing

% Author Orchid ID: enter ID or remove command
\newcommand{\xx}{\boldsymbol{\theta}}
\newcommand{\dx}{d\boldsymbol{\theta}}
\newcommand{\data}{D}
\newcommand{\II}{I}
\newcommand{\mnras}{Monthly Notices of the Royal Astronomical Society}

@@

\section{Algorithm requirements}
The prior $\pi(\xx)$ implies a certain prior for $L_1$ and $L_2$, which we
denote $\pi(L_1, L_2)$, overloading the $\pi$ symbol.
An example of a prior is shown as the
probability density in Figure~\ref{fig:joint1}. The partition function
$Z(\beta_1, \beta_2)$ is a set of expected values with respect to this density.
By analogy with the standard case of one objective function,
simple Monte Carlo sampling from $\pi$ will not work except for values of
$(\beta_1, \beta_2)$ where the canonical distributions of interest are
not too different from $\pi$.
We need a sampler that explores regions where $L_1$ and $L_2$ are increased
(relative to what's typical of $\pi$) in order
to accurately estimate $Z(\beta_1, \beta_2)$. The specific way of increasing
$L_1$ and $L_2$ should follow the requirements listed in the following
subsection, which are unique to NS.

@@

\subsection{Unique Properties of Nested Sampling}
What is special about the NS sequence of distributions (the constrained priors)?
Intuitively, the fact they are chosen adaptively (i.e., the likelihood values
observed
at one iteration determine the likelihood threshold for the next constrained
distribution)
provides a practical advantage
over annealing-based methods (and reduces the number of tuning parameters
substantially --- no `annealing schedule' is needed). While adaptive methods
for generating annealing schedules exist \citep{salomone2018unbiased}, these
are of no use in some phase change problems where a single temperature implies
a mixture of two or more phases.

\subsection{Desiderata}
While we could not see how to satisfy the property in the above subsection,
the algorithm of this paper does satisfy the following other NS properties:
\begin{enumerate}
\item It begins with $N$ points drawn from the prior $\pi(\xx)$.
\item It seeks to explore regions where the values of
$L_1(\xx)$ and $L_2(\xx)$ are higher than the prior $\pi(L_1, L_2)$
would typically imply.
\item It considers a sequence of probability
distributions proportional to $\pi$, but restricted to smaller and smaller
domains for which the enclosed prior mass shrinks by a (roughly) constant and
known factor at each iteration.
\item It is invariant to monotonic transformations of
$L_1$ and $L_2$, i.e. it should only depend on rankings of $L_1$ and $L_2$
values and not the values themselves.
\end{enumerate}
The standard Nested Sampling algorithm has these properties but for a
single objective function. The same is true of variants such as
Diffusive Nested Sampling \citep{dns, dnest4} and PolyChord
\citep{handley2015polychord}.

Sequential Monte Carlo \citep{doucet2001introduction, salomone2018unbiased}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% optional
\appendixtitles{yes} %Leave argument "no" if all appendix headings stay EMPTY (then no dot is printed after "Appendix A"). If the appendix sections contain a heading then change the argument to "yes".
\appendixsections{one} %Leave argument "multiple" if there are multiple sections. Then a counter is printed ("Appendix A"). If there is only one appendix section then change the argument to "one" and no counter is printed ("Appendix").
\appendix
\section{Using the TwinPeaks2018 software}

>>>>

% Author Orchid ID: enter ID or remove command
\newcommand{\xx}{\boldsymbol{\theta}}
\newcommand{\dx}{d\boldsymbol{\theta}}
\newcommand{\data}{D}
\newcommand{\II}{I}
\newcommand{\mnras}{Monthly Notices of the Royal Astronomical Society}
\newcommand{\todo}{\color{blue}{\bf TODO}:~}

@@

\section{Algorithm requirements}
The prior $\pi(\xx)$ implies a certain prior for $L_1$ and $L_2$, which we
denote $\pi(L_1, L_2)$, overloading the $\pi$ symbol.
An example of a prior is shown as the
probability density in Figure~\ref{fig:joint1}. The partition function
$Z(\beta_1, \beta_2)$ is a set of expected values with respect to this density.
By analogy with the standard case of one objective function,
simple Monte Carlo sampling from $\pi$ will not work except for values of
$(\beta_1, \beta_2)$ where the canonical distributions of interest are
not too different from $\pi$.
We need a sampler that explores regions where $L_1$ and $L_2$ are increased
(relative to what's typical of $\pi$) in order
to accurately estimate $Z(\beta_1, \beta_2)$. The specific way of increasing
$L_1$ and $L_2$ should follow the requirements listed in the following
subsection, which are unique to NS. Another unique property of NS which our
algorithm {\em does not} satisfy is discussed in Appendix~\ref{sec:property}.

@@

\subsection{Desiderata}
While we could not see how to satisfy the property in
Appendix~\ref{sec:property},
the algorithm of this paper does satisfy the following other NS properties:
\begin{enumerate}
\item It begins with $N$ points drawn from the prior $\pi(\xx)$.
\item It seeks to explore regions where the values of
$L_1(\xx)$ and $L_2(\xx)$ are higher than the prior $\pi(L_1, L_2)$
would typically imply.
\item It considers a sequence of probability
distributions proportional to $\pi$, but restricted to smaller and smaller
domains for which the enclosed prior mass shrinks by a (roughly) constant and
known factor at each iteration.
\item The exploration procedure is invariant under monotonic transformations of
$L_1$ and $L_2$, i.e. it should only depend on rankings of $L_1$ and $L_2$
values and not on the numeric values themselves.
\end{enumerate}
The standard Nested Sampling algorithm has these properties but for a
single objective function. The same is true of variants such as
Diffusive Nested Sampling \citep{dns, dnest4} and PolyChord
\citep{handley2015polychord}.

\subsection{The algorithm}
The algorithm itself is rather simple. It consists of a collection of
{\em reps} (short for repetitions) of Nested Sampling, each of which ascends
a different compromise between $L_1$ and $L_2$. The output of the reps can then
be used, with appropriate weights, to represent canonical distributions at
any temperature.

The combination of the reps makes use of the (approximate?) unbiasedness
property of Nested Sampling when the prior weights of the dead particles are
assigned proportional to
$[(N-1)/N]^i$ at iteration $i$, instead of being based on
Skilling's geometric sequence of $X$-values given by $\exp(-i/N)$.
\citet{walter} demonstrated that the Nested Sampling estimator of $Z$
is unbiased with these weights, but assuming perfect sampling of the replacement
particles at each iteration. Additionally, using the framework of
Sequential Monte Carlo \citep{doucet2001introduction},
\citet{salomone2018unbiased} showed that this remains true even if imperfect
MCMC is used to generate new particles. Technically, this property is
lost if the sequence of distributions is chosen adaptively (a key feature of NS)
rather than being fixed in advanced. However, this paper demonstrates that the
algorithm works well in practice despite this limitation.

The fact that $Z$ estimates can be made unbiased means that separate $Z$
estimates can be averaged together and the properties of the averaged estimator
will be easily understood and satisfactory ({\todo probably need to be more
technical here.}). {\todo The following statement is based on intuition}
Runs that ascend the `wrong' objective function are still unbiased
estimates of $Z$ for the `right' objective function. Some of the reps will
be well-positioned to represent some canonical distributions but inappropriate
for others. When this happens, the particles from the rep will just have very
low weights with respect to some canonical distributions.

{\todo A description of the algorithm for a single rep follows}.
A probability distribution over the objective functions is set up. If there
are only two objective functions, there is just a probability $p_1$ for
objective function 1, and $1-p_1$ for objective function 2.

The first distribution is the prior $\pi(\xx)$. Objective function thresholds
$(\ell_1, \ell_2)_0 = (-\inf, -\inf)$ are assigned.
$N$ particles are generated
from the prior. Then, an objective function is selected according to the
probabilities $\{p_k\}$. Suppose the objective function chosen is number $k$.
Then the worst particle wrt that criteron is written to disk and replaced,
and the threshold is updated in the $k$th position only.

Essentially, a `rep' is just Nested Sampling, but where a threshold is
maintained for all objective functions. At each iteration, the space is
restricted by only one of the objective functions.

The probability distribution over the objective functions influences the
proportion of the compressions that are done using each objective function.
Different reps ought to have different values for the probability distribution.
Some reps will mostly increase $L_1$, some will mostly increase $L_2$, and
some will be in between.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% optional
\appendixtitles{yes} %Leave argument "no" if all appendix headings stay EMPTY (then no dot is printed after "Appendix A"). If the appendix sections contain a heading then change the argument to "yes".
\appendixsections{multiple} %Leave argument "multiple" if there are multiple sections. Then a counter is printed ("Appendix A"). If there is only one appendix section then change the argument to "one" and no counter is printed ("Appendix").
\appendix
\section{A Unique Property of Nested Sampling}\label{sec:property}
What is special about the NS sequence of distributions (the constrained priors)?
Intuitively, the fact they are chosen adaptively (i.e., the likelihood values
observed
at one iteration determine the likelihood threshold for the next constrained
distribution)
provides a practical advantage
over annealing-based methods (and reduces the number of tuning parameters
substantially --- no `annealing schedule' is needed). While adaptive methods
for generating annealing schedules exist \citep{salomone2018unbiased}, these
are of no use in some phase change problems where a single temperature implies
a mixture of two or more phases.

\section{Using the TwinPeaks2018 software}
